\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-21 paper.qmd
Added trimesh screenshots for PBMC data

Once setting up the hexagon grid with an appropriate number of bins, some hexagon bins may have few or no data points within them (see @fig-bins-scurve (b)). To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagon bins with a considerable number of data points. This involves calculating the number of points within each hexagon. Then, the standard count is computed by dividing the number of points within each hexagon by the maximum number of points in the grid. Next, bins with a standard count less than a benchmark value are removed (see @fig-lwd-scurve (a)). There is no specific rule for selecting a benchmark value. However, the following steps can help determine a suitable value for removing low-density hexagons:

1. Plot the distribution of the standardized counts (see @fig-stdcts-scurve).
2. Examine the distribution of counts.
3. Select the first quantile value if the distribution is skewed.

@@

```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 12

@@

```{r}
#| echo: false
#| fig-cap: "Distribution of standardize counts by hexagons."
#| label: fig-stdcts-scurve
#| out-height: 50%
#| fig-pos: H
#| eval: false

cell_count_scurve
```

@@

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent bechmark choices. What is the effective bechmark value to remove low density hexagons? The residual plot have a steep slope at the end, indicating that a smaller bechmark value causes a small amount of MSE. Then, the slope gradually increases or level up, indicating that a higher bechmark value generates a higher MSE. Using the elbow method, it was observed that when the bechmark value is set to $0.242$, the lowest MSE occurred."
#| label: fig-mse-scurve-lwd
#| fig-pos: H
#| out-height: 50%
#| eval: false

mse_scurve_lwd
```

@@

<!--
```{r}
#| echo: false
#| label: fig-lwd-scurve
#| fig-pos: H
#| fig-cap: "Two approaches to identify low-density hexagons. Hexagon grid (a) marks hexagons with a standardized count less than $0.25$. Hexagon grid (b) marks hexagons by considering the densities of neighboring bins. Removing hexagons by also investigating the densities of neighboring bins is more reliable for preserving the structure."
#| fig-width: 9
#| fig-height: 6

hex_grid_rm_lwd_scurve1 + hex_grid_rm_lwd_scurve2 +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

@@

To create a smooth \gD{} representation (see @fig-NLDR-scurve (d)), it is necessary to remove edges that connect distant bin centroids in the triangular mesh. These edges only exist in the \gD{} model and do not extend into \pD{}, so their removal does not impact the model in \pD{}. Although there are no specific criteria for determining the benchmark value to remove long edges, the following steps provide an approach to identifying a suitable threshold:

1. Plot the distribution of the 2D Euclidean distances (see @fig-dist-scurve).
2. Identify the first largest difference between consecutive distance values.
3. Take the distance value corresponding to this difference as the benchmark value.

@@

Model in \gD{}, on the TriMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape, but it has twists. (The **langevitour** software is used to view the data with a tour, and the full video is available at (<https://youtu.be/zEz2MUCtADM>). 
:::

@@

::: {#fig-pbmc1-sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_1.png)

@@

Screen shots of the **langevitour** of the PBMC3k data set, shows the model-in-data space, a video of the tour animation is available at (<https://youtu.be/VqqWuE0Jj6A>).
:::

In order to find a reasonable NLDR representation for the PBMC3k dataset, the absolute error for different numbers of non-empty bins using various NLDR techniques and different (hyper-)parameter settings (@fig-pbmc-abserror) were calculated. After analyzing the results, it was found that tSNE with default (hyper-)parameter setting (perplexity: $30$) achieved the lowest error when there were $124$ number of non-empty bins. Therefore, tSNE with a perplexity value set to $30$, the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset.

@@

error_plot_pbmc <- ggplot(error_pbmc, 
                              aes(x = b_non_empty, 
                                  y = log(Error), 
                                  colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  geom_vline(xintercept = 136, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")+
    geom_segment(data = arrow_df_pbmc, aes(
      x = x,  
      y = y,  
      xend = xend, 
      yend = yend), 
      arrow = arrow(length = unit(0.01, "npc"),
        ends = "first"), 
        color = "black") +
    inset_element(p = tsne_5_img, 
                  left = 1.7, 
                  bottom = 0.16, 
                  right = 0, 
                  top = 0.31) +                   
    inset_element(p = phate_5_img, 
                  left = 0.6, 
                  bottom = 0.5, 
                  right = 0, 
                  top = 0.65) +                   
    inset_element(p = umap_img, 
                  left = 1.3, 
                  bottom = 0.3, 
                  right = 0, 
                  top = 0.5) +
    inset_element(p = tsne_30_img, 
                  left = 0.6, 
                  bottom = 0.03, 
                  right = 0, 
                  top = 0.23) +
  theme(#aspect.ratio = 0.75,
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90),
        legend.position = "none",
        legend.text = element_text(size = 7))

@@

<!--tSNE with perplexity:30-->
```{r}
#| echo: false
#| fig-cap: "$2\\text{-}D$ layout from tSNE applied for the PBMC3k dataset.  Is this a best representation of the original data? The parameter setting is perplexity=30."
#| label: fig-tsne-suggest
#| fig-pos: H
#| out-width: 40%

@@

::: {#fig-pbmc2-sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_4.png)

@@

Screen shots of the **langevitour** of the PBMC3k data set, shows the model in high-D, a video of the tour animation is available at (<https://youtu.be/5Y1hE4i7N2k>).
:::

@@

Screen shots of the **langevitour** of the MNIST digit 1 data set, shows the model-in-data space, a video of the tour animation is available at (<https://youtu.be/zcg_GXBmqjA>).
:::
<!-- need to update the youtube recording-->

>>>>

Once setting up the hexagon grid with an appropriate number of bins, some hexagon bins may have few or no data points within them (@fig-bins-scurve (b)). To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagon bins with a considerable number of data points. This involves calculating the number of points within each hexagon. Then, the standard count is computed by dividing the number of points within each hexagon by the maximum number of points in the grid. Next, bins with a standard count less than a benchmark value are removed (@fig-param-scurve (c)). There is no specific rule for selecting a benchmark value. However, the following steps can help determine a suitable value for removing low-density hexagons:

1. Plot the distribution of the standardized counts (@fig-param-scurve (b)).
2. Examine the distribution of counts.
3. Select the first quantile value if the distribution is skewed.

@@

```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 11

@@

@@

@@

@@

To create a smooth \gD{} representation (@fig-NLDR-scurve (d)), it is necessary to remove edges that connect distant bin centroids in the triangular mesh. These edges only exist in the \gD{} model and do not extend into \pD{}, so their removal does not impact the model in \pD{}. Although there are no specific criteria for determining the benchmark value to remove long edges, the following steps provide an approach to identifying a suitable threshold:

1. Plot the distribution of the 2D Euclidean distances (@fig-param-scurve (d)).
2. Identify the first largest difference between consecutive distance values.
3. Take the distance value corresponding to this difference as the benchmark value.

@@

Model in \gD{}, on the TriMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape, but it has twists. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/zEz2MUCtADM>). 
:::

@@

::: {#fig-pbmc1-sc layout-ncol="4" fig-pos="H"}
![](figures/pbmc3k/umap_trimesh_plot.png)

![](figures/pbmc3k/sc_1.png)

@@

Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $9\text{-}D$, for the PBMC3k data. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/VqqWuE0Jj6A>).
:::

In order to find a reasonable NLDR representation for the PBMC3k dataset, the absolute error for different numbers of non-empty bins using various NLDR techniques and different (hyper-)parameter settings (@fig-pbmc-abserror) were calculated. After analyzing the results, it was found that tSNE with default (hyper-)parameter setting (perplexity: $30$) achieved the lowest error when there were $136$ number of non-empty bins. Therefore, tSNE with a perplexity value set to $30$, the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset.

@@

error_plot_pbmc <- ggplot(error_pbmc, 
                              aes(x = b_non_empty, 
                                  y = log(Error), 
                                  colour = method)) + 
  geom_point(size = 0.8) +
  geom_line(linewidth = 0.3) + 
  geom_vline(xintercept = 135, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#ff7f00',
                              '#984ea3','#999999','#a65628','#f781bf')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")+
    geom_segment(data = arrow_df_pbmc, aes(
      x = x,  
      y = y,  
      xend = xend, 
      yend = yend), 
      arrow = arrow(length = unit(0.01, "npc"),
        ends = "first"), 
        color = "black") +
    inset_element(p = tsne_5_img, 
                  left = 1.7, 
                  bottom = 0.16, 
                  right = 0, 
                  top = 0.31) +                   
    inset_element(p = phate_5_img, 
                  left = 0.6, 
                  bottom = 0.5, 
                  right = 0, 
                  top = 0.65) +                   
    inset_element(p = umap_img, 
                  left = 1.3, 
                  bottom = 0.3, 
                  right = 0, 
                  top = 0.5) +
    inset_element(p = tsne_30_img, 
                  left = 0.6, 
                  bottom = 0.03, 
                  right = 0, 
                  top = 0.23) +
  theme(#aspect.ratio = 0.75,
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90),
        legend.position = "none",
        legend.text = element_text(size = 7))

@@

<!--tSNE with perplexity:30-->
```{r}
#| echo: false
#| fig-cap: "$2\\text{-}D$ layout from tSNE applied for the PBMC3k dataset.  Is this a best representation of the original data? The parameter setting is perplexity: 30."
#| label: fig-tsne-suggest
#| fig-pos: H
#| out-width: 40%

@@

::: {#fig-pbmc2-sc layout-ncol="4" fig-pos="H"}
![](figures/pbmc3k/tsne_trimesh_plot.png)

![](figures/pbmc3k/sc_4.png)

@@

Model in \gD{}, on the tSNE layout, and three views of the fit in projections from $9\text{-}D$, for the PBMC3k data. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/5Y1hE4i7N2k>).
:::

@@

Model in \gD{}, on the PaCMAP layout, and three views of the fit in projections from $10\text{-}D$, for the digit 1 of MNIST data. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/zcg_GXBmqjA>).

:::
<!-- need to update the youtube recording-->

### 2024-07-21 paper.tex
Added trimesh screenshots for PBMC data

Once setting up the hexagon grid with an appropriate number of bins,
some hexagon bins may have few or no data points within them (see
Figure~\ref{fig-bins-scurve} (b)). To ensure comprehensive coverage of
the NLDR data, it is necessary to select hexagon bins with a
considerable number of data points. This involves calculating the number
of points within each hexagon. Then, the standard count is computed by
dividing the number of points within each hexagon by the maximum number
of points in the grid. Next, bins with a standard count less than a
benchmark value are removed (see \textbf{?@fig-lwd-scurve} (a)). There
is no specific rule for selecting a benchmark value. However, the
following steps can help determine a suitable value for removing
low-density hexagons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the standardized counts (see
  \textbf{?@fig-stdcts-scurve}).
\item
  Examine the distribution of counts.
\item
  Select the first quantile value if the distribution is skewed.
\end{enumerate}

@@

To create a smooth \gD{} representation (see
Figure~\ref{fig-NLDR-scurve} (d)), it is necessary to remove edges that
connect distant bin centroids in the triangular mesh. These edges only
exist in the \gD{} model and do not extend into \pD{}, so their removal
does not impact the model in \pD{}. Although there are no specific
criteria for determining the benchmark value to remove long edges, the
following steps provide an approach to identifying a suitable threshold:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the 2D Euclidean distances (see
  \textbf{?@fig-dist-scurve}).
\item
  Identify the first largest difference between consecutive distance
  values.
\item
  Take the distance value corresponding to this difference as the
  benchmark value.
\end{enumerate}

@@

\caption{\label{fig-scurve-sc-best}Model in \gD{}, on the TriMAP layout,
and three views of the fit in projections from \(7\text{-}D\), for the
S-curve data. The model closely fits the shape, but it has twists. (The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at (\url{https://youtu.be/zEz2MUCtADM}).}

@@

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_1.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_2.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_3.png}\end{minipage}%

\caption{\label{fig-pbmc1-sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model-in-data space, a video of the tour
animation is available at (\url{https://youtu.be/VqqWuE0Jj6A}).}

@@

In order to find a reasonable NLDR representation for the PBMC3k
dataset, the absolute error for different numbers of non-empty bins
using various NLDR techniques and different (hyper-)parameter settings
(Figure~\ref{fig-pbmc-abserror}) were calculated. After analyzing the
results, it was found that tSNE with default (hyper-)parameter setting
(perplexity: \(30\)) achieved the lowest error when there were \(124\)
number of non-empty bins. Therefore, tSNE with a perplexity value set to
\(30\), the default parameter setting, is considered as a reasonable
representation for the PBMC3k dataset.

@@

\caption{\label{fig-tsne-suggest}\(2\text{-}D\) layout from tSNE applied
for the PBMC3k dataset. Is this a best representation of the original
data? The parameter setting is perplexity=30.}

@@

\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_4.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_5.png}\end{minipage}%
%
\begin{minipage}{0.33\linewidth}
\includegraphics{figures/pbmc3k/sc_6.png}\end{minipage}%

\caption{\label{fig-pbmc2-sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/5Y1hE4i7N2k}).}

@@

\caption{\label{fig-mnist1-sc}Screen shots of the \textbf{langevitour}
of the MNIST digit 1 data set, shows the model-in-data space, a video of
the tour animation is available at
(\url{https://youtu.be/zcg_GXBmqjA}).}

>>>>

Once setting up the hexagon grid with an appropriate number of bins,
some hexagon bins may have few or no data points within them
(Figure~\ref{fig-bins-scurve} (b)). To ensure comprehensive coverage of
the NLDR data, it is necessary to select hexagon bins with a
considerable number of data points. This involves calculating the number
of points within each hexagon. Then, the standard count is computed by
dividing the number of points within each hexagon by the maximum number
of points in the grid. Next, bins with a standard count less than a
benchmark value are removed (Figure~\ref{fig-param-scurve} (c)). There
is no specific rule for selecting a benchmark value. However, the
following steps can help determine a suitable value for removing
low-density hexagons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the standardized counts
  (Figure~\ref{fig-param-scurve} (b)).
\item
  Examine the distribution of counts.
\item
  Select the first quantile value if the distribution is skewed.
\end{enumerate}

@@

To create a smooth \gD{} representation (Figure~\ref{fig-NLDR-scurve}
(d)), it is necessary to remove edges that connect distant bin centroids
in the triangular mesh. These edges only exist in the \gD{} model and do
not extend into \pD{}, so their removal does not impact the model in
\pD{}. Although there are no specific criteria for determining the
benchmark value to remove long edges, the following steps provide an
approach to identifying a suitable threshold:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the 2D Euclidean distances
  (Figure~\ref{fig-param-scurve} (d)).
\item
  Identify the first largest difference between consecutive distance
  values.
\item
  Take the distance value corresponding to this difference as the
  benchmark value.
\end{enumerate}

@@

\caption{\label{fig-scurve-sc-best}Model in \gD{}, on the TriMAP layout,
and three views of the fit in projections from \(7\text{-}D\), for the
S-curve data. The model closely fits the shape, but it has twists. (The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at \url{https://youtu.be/zEz2MUCtADM}).}

@@

\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/umap_trimesh_plot.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_1.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_2.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_3.png}\end{minipage}%

\caption{\label{fig-pbmc1-sc}Model in \gD{}, on the UMAP layout, and
three views of the fit in projections from \(9\text{-}D\), for the
PBMC3k data. (The \textbf{langevitour} software is used to view the data
with a tour, and the full video is available at
\url{https://youtu.be/VqqWuE0Jj6A}).}

@@

In order to find a reasonable NLDR representation for the PBMC3k
dataset, the absolute error for different numbers of non-empty bins
using various NLDR techniques and different (hyper-)parameter settings
(Figure~\ref{fig-pbmc-abserror}) were calculated. After analyzing the
results, it was found that tSNE with default (hyper-)parameter setting
(perplexity: \(30\)) achieved the lowest error when there were \(136\)
number of non-empty bins. Therefore, tSNE with a perplexity value set to
\(30\), the default parameter setting, is considered as a reasonable
representation for the PBMC3k dataset.

@@

\caption{\label{fig-tsne-suggest}\(2\text{-}D\) layout from tSNE applied
for the PBMC3k dataset. Is this a best representation of the original
data? The parameter setting is perplexity: 30.}

@@

\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/tsne_trimesh_plot.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_4.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_5.png}\end{minipage}%
%
\begin{minipage}{0.25\linewidth}
\includegraphics{figures/pbmc3k/sc_6.png}\end{minipage}%

\caption{\label{fig-pbmc2-sc}Model in \gD{}, on the tSNE layout, and
three views of the fit in projections from \(9\text{-}D\), for the
PBMC3k data. (The \textbf{langevitour} software is used to view the data
with a tour, and the full video is available at
\url{https://youtu.be/5Y1hE4i7N2k}).}

@@

\caption{\label{fig-mnist1-sc}Model in \gD{}, on the PaCMAP layout, and
three views of the fit in projections from \(10\text{-}D\), for the
digit 1 of MNIST data. (The \textbf{langevitour} software is used to
view the data with a tour, and the full video is available at
\url{https://youtu.be/zcg_GXBmqjA}).}
